---
title: "Practical 12: Planetary Energy Balance"
author: "Thomas Holland"
date: 4th November 2022
output: pdf_document
---
```{r setup, include=FALSE}
# Turn off the r output
knitr::opts_chunk$set(echo = FALSE)

# use mhchem for chemical equations
library(knitr)

# use tufte for formatting
library(tufte)
```

```{r echo=TRUE}
sboltz <- 5.67e-8 # Stefan-Boltzmann constant
wein <- 2.898e-6

dart.temperature = 4500 # K

athos.distance <- 100
athos.solar_flux_density <- 2500
athos.albedo <- 0.2
athos.emissivity <- 0.2

porthos.distance <- 150
porthos.albedo <- 0.3
porthos.emissivity <- 0.5

aramis.distance <- 200
aramis.albedo <- 0.4
aramis.emissivity <- 0.9
```

## Solar output
```{r}
dart.solar_output <- athos.solar_flux_density * athos.distance^2
```

Solar output  $=`r dart.solar_output`$ W

## Planetary solar flux

```{r}
porthos.solar_flux <- dart.solar_output / porthos.distance^2

aramis.solar_flux <- dart.solar_output / aramis.distance^2
```

Porthos solar flux (4 sig fig) $=`r round(porthos.solar_flux, 0)`$ W m$^{-2}$

Aramis solar flux $=`r aramis.solar_flux`$ W m$^{-2}$

## Black body temperature (Kelvin)

```{r}
athos.temperature.bb <- (athos.solar_flux_density * (1-athos.albedo)/4 / sboltz)^(1/4)

porthos.temperature.bb <- (porthos.solar_flux * (1-porthos.albedo)/4 / sboltz)^(1/4)

aramis.temperature.bb <- (aramis.solar_flux * (1-aramis.albedo)/4 / sboltz)^(1/4)
```

Athos temperature $=`r athos.temperature.bb`$ K

Porthos temperature $=`r porthos.temperature.bb`$ K

Aramis temperature $=`r aramis.temperature.bb`$ K

## Black body temperature (Celsius)

```{r}
athos.temperature.bb_c <- athos.temperature.bb - 273.15

porthos.temperature.bb_c <- porthos.temperature.bb - 273.15

aramis.temperature.bb_c <- aramis.temperature.bb - 273.15
```

Athos temperature $=`r athos.temperature.bb_c`$ $^{\circ}$C

Porthos temperature $=`r porthos.temperature.bb_c`$ $^{\circ}$C

Aramis temperature $=`r aramis.temperature.bb_c`$ $^{\circ}$C

## Predicted planetary temperature (Kelvin)

```{r}
athos.temperature.obs <- athos.temperature.bb * (athos.emissivity)^(1/4)

porthos.temperature.obs <- porthos.temperature.bb * (porthos.emissivity)^(1/4)

aramis.temperature.obs <- aramis.temperature.bb * (aramis.emissivity)^(1/4)
```

Athos observed temperature $=`r athos.temperature.obs`$ K

Porthos observed temperature $=`r porthos.temperature.obs`$ K

Aramis observed temperature $=`r aramis.temperature.obs`$ K

## Predicted planetary temperature (Celsius)

```{r}
athos.temperature.obs_c <- athos.temperature.obs - 273.15

porthos.temperature.obs_c <- porthos.temperature.obs - 273.15

aramis.temperature.obs_c <- aramis.temperature.obs - 273.15
```

Athos observed temperature $=`r athos.temperature.obs_c`$ $^{\circ}$C

Porthos observed temperature $=`r porthos.temperature.obs_c`$ $^{\circ}$C

Aramis observed temperature $=`r aramis.temperature.obs_c`$ $^{\circ}$C

## Maximum emission wavelength

```{r}
athos.lambda.max <- wein / athos.temperature.bb

porthos.lambda.max <- wein / porthos.temperature.bb

aramis.lambda.max <- wein / aramis.temperature.bb
```

Athos maximum emission wavelength $=`r athos.lambda.max`$ m

Porthos maximum emission wavelength $=`r porthos.lambda.max`$ m

Aramis maximum emission wavelength $=`r aramis.lambda.max`$ m

## Solar maximum emission wavelength

```{r} 

dart.lambda.max <- wein / dart.temperature
```

D'art solar maximum emission wavelength $=`r dart.lambda.max`$ m

## Green house effect (Kelvin)

```{r}
athos.temperature.gh <- athos.temperature.bb - athos.temperature.obs

porthos.temperature.gh <- porthos.temperature.bb - porthos.temperature.obs

aramis.temperature.gh <- aramis.temperature.bb - aramis.temperature.obs
```

Athos greenhouse effect $=`r athos.temperature.gh`$ K

Porthos greenhouse effect $=`r porthos.temperature.gh`$ K

Aramis greenhouse effect $=`r aramis.temperature.gh`$ K

## Inference from the greenhouse effect

The higher greenhouse effect for Athos suggests that the atmosphere is denser than the other planets.

## Plot of blackbody temperature and mean temperatures (y) against distance (x)

```{r}
x <- c(athos.distance, porthos.distance, aramis.distance)
y <- c(athos.temperature.bb, porthos.temperature.bb, aramis.temperature.bb)
y2 <- c(athos.temperature.obs, porthos.temperature.obs, aramis.temperature.obs)
plot(x, y, type="b", col="red", xlab="Distance (m)", ylab="Temperature (K)", main="Blackbody temperature")
#plot points for x and y2
points(x, y2, col="blue")
lines(x, y2, col="blue")
legend("topright", legend=c("Blackbody", "Observed"), col=c("red", "blue"), lty=1:1)
# range of y axis 190 to 350
axis(2, at=seq(190, 350, by=10))
```



