---
title: "R Markdown"
author: "Thomas Holland (th675)"
date: "December 22, 2022"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# libraries used can be installed using the following commands
# install.packages("tidyverse")
# install.packages("rmarkdown")

# load libraries
library(tidyverse)
```

# Inital data import

## Import data

```{r}
# load data from data.csv using read_csv from the tidyverse package
data <- read_csv("data.csv")

# head of data to check it loaded correctly

head(data)
```

## Data cleaning

```{r}
# print number of rows with missing values
sum(is.na(data))
```

## Log transformation of size data

```{r}
# add new column to data with natural log transformed length

data <- data %>% 
    mutate(log_length = log(length))

head(data)

```

# Basic examination of data

## Box plot of data

```{r}
# box plot of length, grouped by sex from the dataset data, using the ggplot2 package. 

ggplot(data, aes(length, sex)) + 
    geom_boxplot() + 
    geom_jitter(width = 0, height = 0.2, alpha = 0.2) +  # added jitter so overlapping points are visible
    theme_bw() +
    labs(title="Box plot of lengths grouped by the sex of the organism", x="Length (mm)" , y="Sex")
```

```{r}
# box plot of lengths, grouped by sex from the dataset data, using the ggplot2 package. Add x scale every 10 mm

ggplot(data, aes(length, sex)) + 
    geom_boxplot() + 
    geom_jitter(width = 0, height = 0.2, alpha = 0.2) +  # added jitter so overlapping points are visible
    theme_bw() +
    coord_trans(x = "log") + # transform the x axis to natural  log scale
    scale_x_continuous(breaks = seq(0, 100, 10)) + # add more values to the x axis
    labs(title="Box plot of log adjusted lengths grouped by the sex of the organism", x="Length (mm), log scale" , y="Sex")
```

# Statistical tests

This section contains the results of the statistical tests carried out on the data. The analysis of these results and comments are in the analysis section.

## Binomial test

Carry out and report the results of a binomial test to evaluate whether the sex ratio of the sample deviates from 1:1.

```{r}
binomial <- binom.test(table(data$sex), p = 0.5, alternative = "two.sided", conf.level = 0.95)

binomial
```

## T-test

Carry out and report the results of a t-test to evaluate whether females and males differ in size.

```{r}
ttest <- t.test(data$length ~ data$sex, alternative = "two.sided", conf.level = 0.95)

ttest
```

### Diagnostic plots for t-test

Use histogram for diagnostic plots for the t-test.

```{r}
ggplot(data = data, aes(x = length)) + 
    geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue") + 
    geom_density(alpha = 0.2, fill = "red") + 
    facet_wrap(~sex, ncol = 1) + 
    theme_bw() +
    labs(title="Histogram of lengths grouped by sex", x="Length (mm)" , y="Density")
```

## Mann-Whitney U test

Non-parametric alternative to the t-test.

```{r}
mannwhitney <- wilcox.test(data$length ~ data$sex, alternative = "two.sided", conf.level = 0.95)

mannwhitney
```

## T-test on log transformed data

```{r}
ttest_log <- t.test(log(data$length) ~ data$sex, alternative = "two.sided", conf.level = 0.95)

ttest_log
```

### Diagnostic plots for t-test on log transformed data

```{r}
ggplot(data = data, aes(x = length)) + 
    geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue") + 
    geom_density(alpha = 0.2, fill = "red") + 
    facet_wrap(~sex, ncol = 1) + 
    theme_bw() +
    coord_trans(x = "log") + # transform the x axis to natural  log scale
    scale_x_continuous(breaks = seq(0, 100, 10)) + # add more values to the x axis
    labs(title="Histogram of lengths grouped by sex", x="Length (mm), log scale" , y="Density")
```

## F test for ANOVA

```{r}
anova <- aov(data$length ~ data$sex)

summary(anova)
```

```{r}
anova_log <- aov(log(data$length) ~ data$sex)

summary(anova_log)
```

# Analysis

The box plot of the data showed that there was a greater spread of data within the male group than the female group (ANOVA test F = `r paste(anova$`F value`, anova$`Pr(>F)`[1])`)